# Exploratory Data Analysis

In the following section we perform an exploratory data analysis to better
understand the makeup of the meta-data related to the patients.

## Sample-level Descriptive Statistics

The patient meta-data was provided as a `SingleCellExperiment` format.
```{r load the experiment annotation data and count (non)AP, message=FALSE, warning=FALSE}
library(dotenv)
load_dot_env(file = ".env")

library(dplyr)
library(SummarizedExperiment)
library(stringr)

# Load data, note that sce_B.rds should be in data/
experiment = readRDS(paste0(Sys.getenv(c("HOME_DIR")), "data/sce_B.rds"),
                     refhook = NULL)
cellUMI2SampleType_raw = colData(experiment) %>% as.data.frame


cellUMI2SampleType = cellUMI2SampleType_raw %>% select(lib_id,
                                                       Sample.Type,
                                                       Diagnosis, HLADR15) %>%
  unique()

#counting the number of AP and nonAP patients
count_nonap <- sum(str_count(cellUMI2SampleType$Sample.Type, "CFSEhi CD19+"))
count_ap <- sum(str_count(cellUMI2SampleType$Sample.Type, "CFSEdim CD19+"))
```

There are `r count_ap` AP patients and `r count_nonap` nonAP patients
in the dataset.

Next the population statistics is visualised. First we set CFSEdim to
be AP and CFSEhi to be nonAP (see introduction for a rationale).

```{r simplify annotation and save it to disk, message=FALSE, warning=FALSE}
library(ggplot2)

# change CFSEdim CD19+ to AP and CFSEhi CD19+ to nonAP
simplified_annotation = cellUMI2SampleType %>%
  mutate(Sample.Type = ifelse(Sample.Type == "CFSEdim CD19+", "AP", "nonAP"))

#saving the annotations so that we don't have to load it several times
save(simplified_annotation, file = "annotation.RData")
```

```{r show table, message=FALSE, warning=FALSE, echo=FALSE}
head(simplified_annotation)
```

Above we see a part of the dataset with the most important features for the
future analysis.

```{r plot cell count per sample, message=FALSE, warning=FALSE}
#visualisation of the patient composition
p <- ggplot(simplified_annotation, aes(x = Sample.Type, fill = Diagnosis)) +
    geom_bar(position="dodge")  +
    ggtitle("Sample Dataset (n = 31)")

p
```

We see that the proportion of samples is  homogeneous with respect to Diagnosis
and treatment combinations in AP and nonAP sample types.

## Cell-level Descriptive Statistics

Next we want to inspect how many cells we have per AP and nonAP sample. This can
be done by looking at the demultiplexed files, which we discuss in the next section.

```{r, eval=FALSE, message=FALSE}
# since we can't upload the entire demultiplexed folder, this chunk is not 
# evaluated but is here for completeness

library(dplyr)
library(dotenv)
load_dot_env(file = ".env")
samples = list.files(paste0(Sys.getenv(c("HOME_DIR")), "data/demultiplexed"))
cells_for_samples <- data.frame()
for(sample in samples){
    sample_list = list.files(paste0(Sys.getenv(c("HOME_DIR")),
                                   "data/demultiplexed/",
                                   sample))
    sample_list_processed = list()
    # Remove file extension
    for(file in sample_list){
        sample_list_processed = append(sub(pattern = "(.*)\\..*$", replacement = "\\1", basename(file)),
               sample_list_processed)
    }
    sample_list_processed = sample_list_processed %>% unlist() %>% as.data.frame()
    sample = strsplit(sample, "[_]+") %>% unlist() %>% .[1]
    sample = sub("B","", sample)
    sample_list_processed$lib_id = sample
    cells_for_samples = rbind(cells_for_samples, sample_list_processed)
}
colnames(cells_for_samples) = c("cell_id", "lib_id")
save(cells_for_samples, file = "cells_for_samples.RData")
```


```{r load the cell counts, count sample type and visualize, message = FALSE, warning=FALSE}
library(reshape)
library(tidyr)

#load the provided data file
load("cells_for_samples.RData")

separate(data = simplified_annotation, col = lib_id, into = c("lib_id", "rest"), extra = "merge") %>%
  select(-rest) -> simplified_annotation

labels <- simplified_annotation$lib_id
countlabel <- t(c(rep(0,31))) %>% as.data.frame()
colnames(countlabel) <- labels
rownames(countlabel) <- "count"

#counting the number of cells per patient ID
for(label in labels){
 countlabel[label] <- sum(str_count(cells_for_samples$lib_id, label))
}

countlabel <- melt(countlabel)

countlabel$label <- simplified_annotation$Sample.Type
colnames(countlabel) <- c("lib_id", "cell_count", "Sample.Type")

#showing the head of the table
head(countlabel)

#counting the number of cells per condition
cells_ap = 0
cells_nonap = 0
for(i in 1:31){
  if(countlabel[i,"Sample.Type"] == "AP"){
    cells_ap = cells_ap + countlabel[i,"cell_count"]
  }
  else{
    cells_nonap = cells_nonap + countlabel[i,"cell_count"]
  }
}

#visualisation of the cell counts per patient
p <- ggplot(countlabel, aes(x = lib_id, y= cell_count, colour = Sample.Type)) +
  geom_point(stat='identity') + theme(axis.text.x = element_text(angle = 90)) +
  xlab("lib_ID") + ylab ("Count") + ggtitle("Cell count per Sample")

p
```

We see that we have cell counts in the range of $\approx$ 1000 up
to $\approx$ 6500 per sample. Over the two conditions AP and nonAP the
distribution is quite similar with `r format(cells_ap, scientific=FALSE)` AP
cells and `r format(cells_nonap, scientific=FALSE)` nonAP cells.

A check for normality is done via a Q-Q plot and a Shapiro-Wilk normality test.
```{r plot histogram, QQ and do test for normality, message = FALSE}
#histogram of count_data
hist(countlabel$cell_count, xlab = "Cell Count per Sample",
     main = "Histogram of Cell Count")

#adding Q-Q plot to check for normality of data
p <- ggplot(countlabel, aes(sample = cell_count)) +
  stat_qq(col = "blue") + stat_qq_line(col = "red" ) +
  ggtitle("Normal Q-Q plot")
p

#doing shapiro wilk test for normality
shapiro.test(countlabel$cell_count)
```

From our Q-Q plot we see that the data is normally distributed and
the Shapiro-Wilk test indicates a support for a 
normal distribution of the count data.

In a last step the proportion of A P cells vs nonAP cells over all samples
is analysed.
```{r plot proportion of (non)AP, message = FALSE}
library(scales)
#showing the proportion of the cumulative sum of AP vs nonAP cells
summary.df <- data.frame()
summary.df <- cbind(c(cells_ap, cells_nonap))
rownames(summary.df) <- c("AP", "nonAP")
summary.df <- melt(summary.df)

colnames(summary.df) <- c("Type", "X2", "value")

p <- ggplot(summary.df, aes(x = "", y = value, fill = Type)) +
  geom_bar(stat = "identity") + xlab ("Cumsum of Cells") +
  ggtitle("Proportion of overall AP vs nonAP cells") + 
   scale_y_continuous(labels = comma)
p
```

Overall we see that the dataset is well-balanced and we have similar amounts of
patients per diagnosis (Healty Donor, RRMS nihil and RRMS NAT) and we have
a little bit more AP than nonAP cells (AP vs nonAP). Furthermore, it is not the
case that one patient accounts for the majority of cells per condition but that
these cells are quite nicely mixed from different samples (see Cell Count per
Sample plot). The Q-Q plot and the Shapiro-Wilk show that our cell-count data is 
more or less normally distributed.

```{r clean up, message=FALSE, warning=FALSE, include=FALSE, include=FALSE}
# clean up to free up memory
rm(list = ls())
```
