# Frequency of clonal expansion

The files `clonotype_sizes.txt`from `summarize` contain a table with the columns clonotype_sizes and clonotype_count. They can be read into R, as shown below,
and used for downstream analysis. 
```{r load summarized data and calculate annotation frequencies, message=FALSE, warning=FALSE}
library(dotenv)
library(dplyr)
library(SummarizedExperiment)
library(reshape2)
library(tidyr)

load("annotation.RData")

# cut the lib id string and keep only the first part
separate(data = simplified_annotation, 
         col = lib_id, into = c("lib_id", "rest"), extra = "merge") %>% 
  select(-rest) -> simplified_annotation
rownames(simplified_annotation) = simplified_annotation$lib_id

load_dot_env(file = ".env")


# Load data, note that summarise_data should be in data/
file_path = paste0(Sys.getenv(c("HOME_DIR")),"data/summarise_data")

# determine the highest clonotype size in the data
max_clonotype_size = 0
for (i in list.files(path = file_path)){

  tmp_path = paste0(file_path,"/",i,"/clonotype_sizes.txt")
  tmp = read.table(tmp_path, header = TRUE)


  if (max(tmp$clonotype_size) > max_clonotype_size){
    max_clonotype_size = max(tmp$clonotype_size)
  }
}

# prepare empty data frame (df for frequency and df_complete for clonotype counts)
df = data.frame(lib_id = character(), expansion_frequency = double())
df_complete = data.frame()
lib_ids = c()

# loop over all files and fill the dataframes
for (i in list.files(path = file_path)){

  tmp_path = paste0(file_path,"/",i,"/clonotype_sizes.txt")
  tmp = read.table(tmp_path, header = TRUE)

  # calculate frequency of clonal expansion (clonotype size >= 2)
  freq=tmp$clonotype_size*tmp$clonotype_count
  sum_freq=sum(freq)
  freq=(sum_freq - freq[1])/sum_freq

  #split the string at "-" or "_" to remove "-preprocessed" and "POOL*" and
  # remove the "B"
  lib_name = strsplit(i, "[-_]+") %>% unlist() %>% .[1]
  lib_name = sub("B","",lib_name)

  expand_tmp = list(lib_id = lib_name, expansion_frequency = freq)
  df = rbind(df,expand_tmp)

  lib_ids = c(lib_ids,lib_name)

  # fill vector with zeros up to max clonotype size to have equal lengths
  adjusted_clonotype_count = c(tmp$clonotype_count,
                              rep(0,(max_clonotype_size -
                              max(tmp$clonotype_size))))

  df_complete = rbind(df_complete, adjusted_clonotype_count)
  }

colnames(df_complete) = 1:max_clonotype_size
rownames(df_complete) = lib_ids
```

```{r tmp table, echo=FALSE}
head(tmp)
```
Shown above is a part of such a table. A clonotype of size 1 means, that only 
one cell is in the cluster and is therefore unique.
The tables of each sample (each folder in the output directory) are combined 
into a dataframe called `df_complete`. The column index of the dataframe
indicates the clonotype size and the rownames specify the lib_id (sample name).
Since the table of each lib_id directory has the same number of rows as the
highest clonotype_size of that lib_id, the missing entries of `df_complete` are 
filled with zeros.
A section of `df_complete` is shown below.
The highest clonotype_size in our data is `r dim(df_complete)[2]`
```{r show df_complete, echo=FALSE}
df_complete[1:3,1:24]
```


We can also calculate the frequency of clonal expansion, using the definiton of 
clonal expansion of [@ramesh]
"as a highly similar B cell cluster containing two or more B cells".
The clonal expansion frequency of a sample is therefore

$$
\frac{\text{number of cells with clonotype_size >1}}{\text{total number of cell in the sample}}
$$

The clonal expansion frequencies of each sample are combined in the
dataframe `df`.
```{r, echo=FALSE}
head(df)
```

For better visualization we also make a data frame where we pool all clonotype
counts of size >4 into one category. We also make a separate data frame where we normalize by the number of cells per sample to get a proportion.
```{r condensing dataframe}
# condense all clonotypes > 4 into one category and make a new data frame
condensed_sum = rowSums(df_complete[,-c(1:4)])
df_summarized = cbind(df_complete[,c(1:4)],condensed_sum)
colnames(df_summarized) = c(1:4, "> 4")

# normalize by rowsum
df_summarized_normalized = df_summarized/rowSums(df_summarized)
head(df_summarized)
```



After we add the annotation data to the dataframes we are ready for plotting.
```{r add the sample type to the dataframe, message=FALSE, warning=FALSE}
library(tidyr)
library(gtable)

# merge the annotation with the data frames
df = merge(df, simplified_annotation, by.x = "lib_id", by.y = "lib_id")
rownames(simplified_annotation) = simplified_annotation$lib_id

df_summarized = merge(df_summarized, simplified_annotation, by = "row.names")

rownames(df_summarized) = df_summarized$Row.names
df_summarized = subset(df_summarized, select = -Row.names)

df_summarized_normalized = merge(df_summarized_normalized, simplified_annotation, by = "row.names")

rownames(df_summarized_normalized) = df_summarized_normalized$Row.names
df_summarized_normalized = subset(df_summarized_normalized, select = -Row.names)
```


If we try to analyse the clonotype vectors directly (so a row in df_complete above)
and look how similar in terms of a euclidean distance they are, we can
perform a hierarchical clustering and get the dendrogram below and perform a PCA


```{r plot dendogramm, message=FALSE, warning=FALSE}
#plotting the clustering of id's according to their clonotype
library(dendextend)
library(ggfortify)

df_cluster = df_summarized[,1:5]

col = df_summarized$Sample.Type
col = ifelse(col=="AP","black","red")

dist_mat = dist(df_cluster, method = 'manhattan')
hclust_avg = hclust(dist_mat, method = 'average')
avg_dend_obj = as.dendrogram(hclust_avg)

colors_to_use <- as.numeric(as.factor(col))
colors_to_use <- colors_to_use[order.dendrogram(avg_dend_obj)]

labels_colors(avg_dend_obj) <- colors_to_use

plot(avg_dend_obj, main = "Dendrogram of Samples coloured by AP vs nonAP")
legend("topright", legend = c("AP", "nonAP"), col = c("black", "red"), pch=c(20,20))
```

```{r plot the contributions of the PCA, message=FALSE, warning=FALSE}
# plot the PCA of the datapoints
pca = prcomp(df_cluster, center = TRUE, scale = TRUE)
plot(pca, main = "Principle Component Analysis", xlab = "PCs")
```
```{r plot the PCA, message=FALSE, warning=FALSE, fig.width=14, fig.height=12}
library(ggrepel)
labels = df$lib_id
outlier_labels = c("IJ020","IJ028", "IJ067")
labels[df$lib_id %in% outlier_labels %>% !.] = ""
autoplot(pca, scale = 0, colour = col, legend = TRUE) +
  geom_text_repel(vjust=1.5, label=labels, min.segment.length = Inf) +
  ggtitle("PCA plot AP (black) vs nonAP (red)")
```


We find that AP and nonAP samples actually cluster together already in terms of
the clonotype sizes. When looking at the PCA, we see that we can explain
most of the variability in our data with two principle components. When plotting
the points, we note that one point (IJ028) prevents our data to be linearly
separable. In the hierarchical clustering we see as well, that IJ028
clusters with nonAP cells as well as the sample IJ020, which is the second
closest point to the putative separation line.

Plotted below are the absoulte counts of clonal expansion grouped by sample type and the proportion of clonal expansion grouped by sample type and diagnosis.
```{r, plot the absolute clonal expansion, fig.width=14}
library(ggplot2)
melted = melt(df_summarized, id.vars = c("lib_id", "Sample.Type", "Diagnosis", "HLADR15"))
ggplot(data = melted, aes(fill=variable, x=lib_id, y=value)) +
  geom_bar(stat="identity", position = position_stack(reverse = TRUE)) +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_grid(.~Sample.Type, scales = "free_x") +
  labs(fill = "Clonal expansion") +
  xlab("Sample Id") + ylab("Clonotype Count")
```

Looking at the plots we see that one point in the AP samples is again a bit 
anomalous, with lesse expansion than the others. It is again IJ028 that we 
saw clusters with the nonAP samples (see above).

```{r, plot the relative clonal expansion, fig.width=14, fig.height=10}
library(ggplot2)
melted = melt(df_summarized_normalized, id.vars = c("lib_id", "Sample.Type", "Diagnosis", "HLADR15"))
ggplot(data = melted, aes(fill=variable, x=lib_id, y=value)) +
  geom_bar(stat="identity", position = position_stack(reverse = TRUE)) +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(Sample.Type~Diagnosis, scales = "free_x") +
  labs(fill = "Clonal expansion \nclonotype size") +
  xlab("Sample Id") + ylab("Proportion of Clonotype Count")
```

If we look at the proportion of clonotype counts and compare the sample types we can see an increase in all clonotype size groups of type `AP`vs type `nonAP`. But we do not see a patten with respect to `diagnosis`.

Next we look at the frequencies of clonal expanion.
```{r plot the frequencies, message=FALSE, warning=FALSE}
library(ggplot2)
p <- ggplot(df, aes(x = Sample.Type, y = expansion_frequency, color = Diagnosis)) + geom_point() +
  xlab("Sample Type") + ylab("clonal expansion frequency") + ylim(0,1) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p
```

Here we again see the increase in the frequency of `AP` vs `nonAP` but no pattern with respect to diagnosis.
There is one AP point that overlaps in its clonal expansion frequency with 
nonAP and upon checking we find again, that it is  our anomalous sample IJ028
that has the lowest frequency of all of the samples.

To verify our findings we will do a Mann–Whitney U test.
```{r do shapiro test}
#Shapiro-Wilk test to assess whether expansion frequencies are normal or not
shapiro.test(df$expansion_frequency)
```

The Shapiro-Wilk test justifies the usage of a nonparameteric test as the
Wilcoxon test used here.

```{r Mann-Whitney-Wilcoxon Test}
wilcox.test(expansion_frequency ~ Sample.Type, data=df)
```
As we can see the null hypothesis of true location shift is equal to 0 can be rejected under the significance level $\alpha = 0.05$.

As an additional visualization we want to see if there is a pattern in the clonotype expansion frequency with regard to the haplotype HLA-DR15, but we again do not see a pattern.
```{r plot frequency for HLADR15}
library(ggplot2)
p <- ggplot(df %>% subset(HLADR15 != "?"), aes(x = Sample.Type, y = expansion_frequency,
                                               color = Diagnosis)) + geom_point() +
  xlab("Sample Type") + ylab("clonal expansion frequency") + ylim(0,1) +
  theme_bw() +
  facet_grid(.~HLADR15, scales = "free_x") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p
```

```{r, include=FALSE}
# clean up to free up memory
rm(list = ls())
```
