# Isotype Analysis 

The files `BCR_summary.txt`from `summarize` contain a section *Isotype of cells with productive heavy chain*. As the name says it counts the isotype of the B-cell receptor of each cell in the sample. The section is a table with the columns `Isotype`, `cells` and `% of cells`. This can be read into R, as shown below, and use for downstream analysis.
```{r Isotype analysis, message = FALSE, warning=FALSE}
library(dplyr)
library(tidyr)

#load annotations
load("annotation.RData")

# cut the lib id string and keep only the first part
separate(data = simplified_annotation, col = lib_id, into = c("lib_id", "rest"), extra = "merge") %>%
  select(-rest) -> simplified_annotation

# Load data, note that summarise_data should be in data/
file_path = paste0(Sys.getenv(c("HOME_DIR")),"data/summarise_data")

# prepare empty df
isotype_df = data.frame()

# loop over all files and fill the dataframes
for (i in list.files(path = file_path)){
  tmp_path = paste0(file_path,"/",i,"/BCR_summary.txt")

  # read the file line by line
  l = readLines(tmp_path)

  # find the start about the section about the isotypes
  start_line = grep("##Isotype of cells with productive heavy chain##",l)+2

  # find empty lines in file
  empty_lines = which(!nzchar(l))

  # find first empty line after starting line
  end_line = empty_lines[empty_lines > start_line][1]-1

  tmp_table = read.table(tmp_path, skip = start_line, nrows =
                          (end_line-start_line), header = FALSE) %>%
    as.data.frame()

  colnames(tmp_table) = c("Isotype", "cells", "percent_of_cells")

  #split the string at "-" or "_" to remove "-preprocessed" and "POOL*"
  # and remove the "B"
  lib_name = strsplit(i, "[-_]+") %>% unlist() %>% .[1]
  lib_name = sub("B","",lib_name)
  
  tmp_table$lib_id = lib_name

  isotype_df = rbind(isotype_df, tmp_table)
}

# merge the annotation with the data frames
isotype_df = merge(isotype_df, simplified_annotation, by.x = "lib_id", by.y = "lib_id")
```

An example of such a table is shown below.
```{r, echo=FALSE}
head(tmp_table[,-4])
```
The tables were combined into a new data frame `isotype_df` and annotated. A section of the data frame is shown below.
```{r, echo=FALSE}
head(isotype_df)
```
We can plot the proportion of the isotype (% of cells) for each isotype as a boxplot and compare the sample types.

```{r calculate comparisons, message=FALSE, warning=FALSE}
library(ggplot2)
library(ggpubr)
library(rstatix)

stat.test <- isotype_df %>%
  group_by(Isotype) %>%
  wilcox_test(percent_of_cells ~ Sample.Type) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj") %>%
  add_xy_position(x = "Isotype") %>%
  mutate(y.position = 100)
```

```{r convert adjusted p value to symbol}
symbols = symnum(stat.test$p.adj, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                        symbols = c("****", "***", "**", "*", "ns"))
stat.test$symbol = symbols
```


```{r isotype boxplots, warning=FALSE, message=FALSE}
p <- ggplot(isotype_df, aes(x = Isotype, y = percent_of_cells, color = Sample.Type)) +
  geom_boxplot()+ xlab("Isotype") + ylab("proportion (%)") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  stat_pvalue_manual(stat.test,  label = "symbol", tip.length = 0) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))+
  ylim(0,100)


p
```

The comparisons were done with the Wilcoxon test, correction for multiple testing was done with Benjamini–Hochberg. The significance is indicated with stars 


| symbol | meaning      |
|--------|--------------|
| ns     | P > 0.05     |
| *      | P $\le$ 0.05 |
| **     | P $\le$ 0.01 |
| ***    | P $\le$ 0.001|


We see that when comparing `AP`with `nonAP` we have a significant increase in *Unknown* isotypes (isotype could not be determined by bracer), although the effect size is small. We also have a increase in isotypes *IGHG2*. We have a significant decrease in isotypes *IGHD*, *IGHDM* and *IGHM*.
